Estimating Daily Bicycle Counts in Seattle from Seasonal and Weather Factors

### 2015-07-30 article/article.tex
format for TRB

\documentclass[12pt,letterpaper,article,twocolumn]{memoir}

% Layout options
\chapterstyle{article}
\setulmarginsandblock{.75in}{1in}{*}
\setlrmarginsandblock{.75in}{*}{*}
\setcolsepandrule{.33in}{0pt}

@@

@@

@@

@@

\maketitle

\section*{Introduction}
%
% 1. Explain what we set out to do (i.e. understand influence of
% seasonal and weather factors on bicycling) 2. explain relevance to
% planners (e.g. understanding how much seasonal variability to expect
% for service provision, understand methodologically how to use data
% collected from counters) 3. ??? other things I'm forgetting right
% now.
%

@@

This article contributes to that understanding by focusing on the
substantive effects of weather and seasonality on the number of
bicyclists observed on a given day. Specifically, we hypothesize that
factors associated with inclement weather (esp. low temperature and
precipitation) will be inversely correlated with bicycle counts. We
similarly expect to see an additional decrease in counts associated
with both physical and social/institutional dimensions of season. That
is, in seasons where the days get shorter, we expect fewer cyclists
due to a kind of decision making inertia. We also expect fewer
cyclists outside the ``school season'' as we believe that a large
number of bicycling trips are associated with commuting to schools
that, unlike most other institutions have clear annual cycles.

@@

\section*{Literature review}
% 
% Identify where people have talked about weather and seasonal factors
% influencing counts previously. Describe their methods together with any
% limitations, as well as their substantive findings---especially what
% variables they considered. Normally should conduct this first, but
% our driver here is a class assignment! We should fill this in and
% adjust our analysis accordingly before attempting to go to press.
%

@@

Research has found the variability for counts has a positive
association with high temperature and low precipitation. \parencite{Niemeier:1996aa,Parkin:2008aa}. Meanwhile, as suggested by
\textcite{Lewin:2011aa} and \textcite{Thomas:2009aa}, the effects of
precipitation and temperature on bicycle volumes are nonlinear. For
example, bicycle traffic can decrease in both very cold and very hot
weather as noted by \textcite{Richardson:2000aa}. Apart from the usual
temperature and rain variables, \textcite{Miranda-Moreno:2011aa} finds
humidity and additional precipitation variables including the presence
of rain in the morning and/or during the previous three hours to be
significant too. Other comparative studies are also available where
bicycle counts are conducted in different cities, and different
sensitives to weather are examined \parencite{Rose:2011aa}. As for
longitudinal studies, \textcite{Niemeier:1996aa} finds increased variability
for counts conducted in the later months of the year.
\textcite{Jones:2010aa} conclude that morning peak hours from 6 AM to 9 PM
accounts for a consistent 95\% of the total bicycle volumes by hourly
count data. The simple linear regression model has been used in
several applications \parencite{Jones:2008aa,Jones:2010aa}. Other modeling approaches include \textcite{Miranda-Moreno:2011aa} which develops a count
model and \textcite{Thomas:2009aa} which develops a time-series model.
\textcite{Niemeier:1996aa} also uses a Poisson model to statistically
confirm  many of the factors thought to influence cyclists. The
work by \textcite{Gallop:2012aa} adopts a similar time-series approach
while incorporating an autoregressive integrated moving average
(ARIMA) analysis. A summary of findings from the literature are
presented in table \ref{tb:lit}.

These literature together suggest an opportunity for further model
development base around long-term automated counts utilizing
appropriate statistical methods. How seasonal factors influence
bicycle flow needs to be examined in data that last more than a year.
One limitation present in much of the past literature is that few
discuss goodness of fit of their modeling. Further a model that can
better describe and forecast the bicycle count in longitudinal form is
necessary to developed. Models for count data with better estimation
methods offer some promise.

\newfixedcaption{\freetabcaption}{table}
\begin{centering}
\begin{scriptsize}
\freetabcaption{Summary of findings from literature review}
\begin{ctabular}{p{.6in}p{1.15in}p{1.15in}}
  \toprule
  Source & Variable(s) identified & Methods \\
  \midrule
  \cite{Fields:2012aa} & Average daily temperature (F); Total weekly precipitation (inches) & Identify patterns through scatter plots; No explicit model is established. \\
  \cite{Gallop:2012aa} & Temperature, Relative humidity, wind speed, clearness,fog, precipitation & Use ARIMA to account for serial correlation patterns \\
  \cite{Griswold:2011aa} & Nearby population and employment density, proximity to downtown/freeway, age, education level, income, et. & Log linear ordinary least squares regression is used to estimate a bicycle count model \\
  \cite{Helbich:2014aa} & Daily maiximum air temperature (C), daily average wind speed (m/s), daily precipitation(mm) & Place-specific associations of weather conditions are explored through geographically weighted logit models \\
  \cite{Hunt:2007aa} & Descriptive variables indicating lane use, secured parking, level of experience, etc. & Logistic model of cycling-related choices \\
  \cite{Jones:2010aa} & Length of bicycle network, employment density, population density & Standard ordinary least squares regression \\
  \cite{Lewin:2011aa} & Max temp (F), rain flag, snow flag, weekend flag, over 90 flag & Standard linear regression model \\
  \cite{McCahil:2008aa} & logarithmic choice measure, population density, worker density & A new space syntax theory is used to evaluate and predict the bicycle volume throughout a network \\
  \cite{Miranda-Moreno:2011aa} & Temperature (C), humidity (\%), rain presence (0/1), rain in prev. 3hrs (0/1), warm \& humid (0/1), morning rain (0/1) & Both log-linear model and negative binomial model are tested \\
  \cite{Niemeier:1996aa} & Morning flag, rain flag, high temp flag, location variable, season variable & A Poisson bicycle count model is assumed and fitted \\
  \cite{Nosal:2014aa} & Temperature (F), humidity (\%), rain flag, rain prev. 3hrs, am rain, pm rain & The relationship  is analyzed using a log-linear regression model \\
  \cite{Parkin:2008aa} & Gender, car ownership, hilliness, off-road routes proportion & A logistic regression model of relevant socio-economic and physical variables is estimated. \\
  \cite{Pinjari:2009aa} & Household density, employment density, fraction of commercial land area, fraction of senior, fraction of race population, etc. & The model system takes the form of a joint mixed Multinomial Logit–Multiple Discrete-Continuous Extreme Value (MNL-MDCEV) structure \\
  \cite{Rose:2011aa} & Temerature(C), rainfall(mm), holiday flag, school seanson flag, day of the week & Weather and other effects examined using an aggregate model of daily ridership \\
  \cite{Thomas:2009aa} & Temperature(C), sunshine(h), precipitation(mm), wind force(m/s), cycle path use & A bi-level structure is developed with the upper level being a log-linear model and the lower level being a linear model \\
  \bottomrule
\label{tb:lit}
\end{ctabular}
\end{scriptsize}
\end{centering}

\section*{Methodology}
In order to discern the relationship between bicycle counts and
several identified weather, seasonal, and temporal factors, we
developed a statistical model that attempts to predict daily bicycle
counts from these other factors. This section describes the methods
and procedures we used to collect and process the estimation dataset,
the rationale for our selection of variables, our chosen model type,
and our model estimation procedures.

\FloatBarrier
\subsection*{Data collection, processing, description}
%
% Data collection: study site, descriptives of sample, collection from
% forecast.io and seattle.gov, not in that order
%

The data that we attempt to explain were collected at Seattle's
Fremont Bridge and cover a period of two years spanning from October
31, 2012 to October 30, 2014. The Fremont Bridge captures a
substantial amount of bicycle traffic due to its status as one of only
five facilities that carry bicyclists across the canal separating the
northern and southern halves of Seattle.

@@

Figures \ref{fg:hist} and \ref{fg:timeseries} provide visual summaries
of the processed counts data. Some apparent outliers are visible at
the rightmost portion of the histogram. The two highest counts
occurred the Monday and Tuesday preceding the beginning of National
Bike to Work Month. And the third highest count occurred on National
Bike to Work Day.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/dp1}
  \caption{Histogram showing frequency of observed bicycles}
  \label{fg:hist}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/dp2}
  \caption{Timeseries plot of bicycle counts}
  \label{fg:timeseries}
\end{figure}

\FloatBarrier
\subsection*{Variable selection}
% 
% Data analysis: specification of model, choice of model (nb v pois),
% description of counterfactual simulation using the estimated model,
% software used.
%

Out of the set independent variables retrieved from Forecast.io and
the University of Washington, we selected a subset that we felt best
reflected our specific research questions. The final set of variables
included in our model include Table \ref{tb:variables} summarizes the
variables used in our model.

\begin{centering}
\begin{table}[ht!]
\begin{scriptsize}
\caption{Variables included in model specification}
\begin{tabularx}{0.47\textwidth}{>{\raggedright\arraybackslash}p{1.6cm}>{\raggedright\arraybackslash}X}
\toprule
Variable & Description \\
\midrule
Count* & Number of bicycles per day \\
Daylight  & Length of daylight in hours \\
UW & The University of Washington was in session (TRUE/FALSE) \\
Holiday & The day was a holiday as recognized by UW (TRUE/FALSE) \\
Max temp & Maximum temperature in Fahrenheit for the day \\
Max precip & Daily maximum inches of precipitation in any hour \\
Sat--Fri & Day of the week dummy variable (TRUE/FALSE, relative to Sunday) \\
Day \# & Sequentially numbered day of study \\
\bottomrule
\multicolumn{2}{r}{* Dependent variable}
\end{tabularx}

\label{tb:variables}
\end{scriptsize}
\end{table}
\end{centering}

@@

Daily maximum temperature, measured in Fahrenheit, was chosen to
represent temperature (rather than, for example, substituting or
adding daily minimum temperature) in part to retain simplicity in the
model, in part because there is relatively little daily temperature
variation in Seattle due to the moderating effect of large water
bodies, and in part because maximum temperature better reflects the
conditions during daylight hours when most bicycle trips would occur.
This simplification may not be warranted for other locations that
experience greater temperature variation than Seattle.

@@

\FloatBarrier
\subsection*{Model estimation and goodness of fit}
% Model choice
A natural model choice for count data is the Poisson model, however we
believed that our count data were overdispersed. Overdispersion, or
contagion between events, violates the mean-variance equivalence
assumption of the Poisson model. While overdispersion would not impact
our parameter estimates, it would result in overly optimistic margins
of error. In order to account for and estimate the amount of
overdispersion present in our data, we chose a negative binomial model
type. Because there were no days observed with zero bicycle counts, we
did not need to resort to zero-inflated models as is often necessary
when dealing with count data.

@@

Additionally, we compared the residuals plots for both the Poisson and
negative binomial models. As seen in figures \ref{fg:resnb} and
\ref{fg:respois}, the residual variance is much lower in the negative
binomial model, further confirming the choice of a negative
binomial for these counts.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/residuals_pois}
  \caption{Residual variance from the Poisson model}
  \label{fg:respois}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/residuals}
  \caption{Residual variance from the negative binomial model}
  \label{fg:resnb}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/avpp2-sm}
  \caption{Actual versus predicted values as fit by negative binomial model}
  \label{fg:avp}
\end{figure}

@@

\section*{Results}
% 
% One subsection for each research question
% 
This section presents the results from the statistical model and
accompanying counterfactual simulations as described in the preceding
section. Each of the main research questions of seasonality, weather,
and general trend are addressed here. In addition, additional added
control variables such as day of the week, holidays, and linear trend
are presented.

Without exception, each of the coefficients in our model was
statistically significant at the $p < 0.05$ level. Further, with the
exception of the Saturday coefficient, all coefficients were
significant at the $p < 0.001$ level. The remainder of this section
focuses on presenting the substantive effect of each variable.

\FloatBarrier
\subsection*{Seasonality}
As discussed previously, we considered two variables to address the
question of seasonality: the first being the number of daylight hours,
and the second being whether or not the University of Washington
(proxying more generally for other educational institutions) was in
session.

@@

\FloatBarrier
\subsection*{Weather}
As in the case of seasonality, our model represents weather using two
variables, in this case precipitation (measured as the maximum amount
of precipitation falling in any hour of that day), and maximum
temperature (measured daily, in Fahrenheit).

@@

Temperature, in contrast to precipitation, has a clearly positive
association with increased numbers of bicyclists. We were somewhat
surprised to not see a leveling off in counts at very high
temperatures---likely due to Seattle's moderate summer climate, and the
lack of extremely warm days in our dataset.

@@

\FloatBarrier
\subsection*{General trend in bicycle counts}
Our results (shown in figure \ref{fg:trend}) confirm the presence of a
general trend toward increased numbers of bicycles at this location.
With all else held constant, we would expect to see roughly 300 more
bicycles on days at the end of our study period than at the beginning.
This is consistent with figures reported elsewhere that suggest that
bicycling is increasing in a number of cities including
Seattle \parencite{League-of-American-Bicyclists:aa}.

@@

\FloatBarrier
\subsection*{Day of week variation}
As discussed previously, day of the week was included due to the
weekly variation in bicycle counts apparent in timeseries plots.
Figure \ref{fg:dayofweek} shows several interesting aspects of these
results. First, we see much higher numbers of bicyclists on weekdays
than on weekends. This strongly suggests that the majority of the
bicycle traffic at this location is for commuter purposes.

@@

\FloatBarrier
\section*{Discussion}
This research set out to better understand the relationship between
bicycle counts and weather and season. However, weather and season are
themselves comprised by a number of constituent elements. We believe
this research has chosen defensible representations of weather and
season. However, we readily concede that there exist other reasonable
representations of season and weather.

@@

One variable missing from our analysis that we would have liked to
include is cloud cover.\footnote{We gratefully acknowledge Christopher
  Adolph for this suggestion.} This variable would have allowed us to
account for the possibility of brilliant, sunny, but cold winter days,
as well as warm, dry, but gloomy summer days. Unfortunately, the
dataset downloaded from Forecast.io included a great deal of missing
values for this variable, and we were unable to identify another
suitable source for this data in time for this article.

Another limitation of this research comes from our sample of bicycle
counts. While we do have continuous counts spanning a full two years,
these counts were taken at just one point location. We believe the
Fremont bridge is somewhat representative of a high-volume bicycle
facility in Seattle, however we might, for example, see a decreased
effect size for University of Washington session status if we looked
at counts further from the university. The City of Seattle has begun
collecting and releasing data from other counters, which could be
included in a future version of this analysis.

Finally, as discussed in the methodology section, we believe the
choice of the negative binomial model to be a better choice than the
Poisson, which had been used in some previous studies. A limitation in
this model, however, is in not accounting for timeseries
autocorrelation. We think it highly likely that the decision to
bicycle on one day might influence the decision to bicycle on the
next. A future direction would be to consider models, such as ARIMA,
that more explicitly account for time.

\section*{Conclusion}
This research set out to help clarify the relationship between weather
and seasonal factors and bicycle counts in Seattle. In order to
achieve this, we developed a negative binomial model to predict
bicycle counts based on temperature, precipitation, day length,
university in-session status, day of the week, and a general linear
time trend.

For each term included in our model, we found statistically
significant effects. More importantly, through the use of
counterfactual simulation, we estimated what we deem to be substantial
effect sizes associated with each predictor variable.

This article contributes to the existing literature by demonstrating
the use of a negative binomial model for bicycle counts. In addition,
the results presented here were generated from data collected over a
long study period of two years.

>>>>

\documentclass[12pt,letterpaper,article]{memoir}

% Layout options
\chapterstyle{article}
\setulmarginsandblock{1in}{1in}{*}
\setlrmarginsandblock{1in}{*}{*}
\setcolsepandrule{.33in}{0pt}

@@

% Font
\usepackage{tgtermes}

% TRB Line numbers
\usepackage{lineno}
\usepackage{fnlineno}
\linenumbers

@@

% Needed for sub-figures
\newsubfloat{figure}

@@

% Header and footer
\makepagestyle{trb}
\makeevenhead{trb}{Schmiedeskamp, Zhao}{}{\thepage}
\makeoddhead{trb}{Schmiedeskamp, Zhao}{}{\thepage}

% Headings styles
\makechapterstyle{trb}{%
\renewcommand{\chapterheadstart}{\vspace*{3pc}}
\renewcommand{\chapnamefont}{\bfseries\raggedright}
\renewcommand{\chapnumfont}{\bfseries\raggedright}
\renewcommand{\chaptitlefont}{\bfseries\raggedright}
\renewcommand{\printchaptername}{\S}
\renewcommand{\chapternamenum}{ }
\renewcommand{\printchapternum}{\chapnumfont \thechapter}
\renewcommand{\afterchapternum}{. }
\renewcommand{\afterchaptertitle}{\par\nobreak\vspace*{1pc}}
\renewcommand*{\printchaptertitle}[1]{\chaptitlefont\MakeUppercase{##1}} 
}
\setsecheadstyle{\bfseries\raggedright}
\setsubsecheadstyle{\itshape\raggedright} % danger zone

% Table and figure captions
% http://tex.stackexchange.com/questions/75654/uppercase-until-end-of-group
\makeatletter
\def\redeffloatcapstyle#1{%
  \expandafter\renewcommand\expandafter{\csname fnum@#1\endcsname}[1]{%
    \bfseries\MakeUppercase{%
      \expandafter{\csname #1name\endcsname}~\expandafter{\csname the#1\endcsname}\enskip
    }
  }
}
\makeatother
\redeffloatcapstyle{figure}
\redeffloatcapstyle{table}
%\captionnamefont{\bfseries\scshape}
\captiontitlefont{\bfseries}
\captiondelim{}

\pagestyle{trb}
\chapterstyle{trb}

@@

\thispagestyle{empty}

\par\noindent\textbf{ESTIMATING DAILY BICYCLE COUNTS IN SEATTLE FROM SEASONAL AND WEATHER FACTORS}

\vspace*{4pc}

\par\noindent\textbf{Peter Schmiedeskamp} (Corresponding author)\\
University of Washington, Department of Urban Design and Planning\\
Box 355740, Seattle, WA 98195-5740\\
Tel: 206.289.0694\\
email: pschmied@uw.edu

\vspace*{1.5pc}
\par\noindent\textbf{Weiran Zhao}\\
University of Washington, Department of Urban Design and Planning\\
Box 355740, Seattle, WA 98195-5740\\
Tel: 206.310.7065\\
email: weiran36@uw.edu

\vspace*{3pc}
\par\noindent Word count: \textbf{7,452} (4,952 text \& bibliography + 8 figures + 2 tables)

\vspace*{3pc}
\par\noindent Submission Date: \textbf{1 August 2015}

\clearpage

\par\noindent\textbf{ABSTRACT}

\par\noindent This paper examines the relationship between several seasonal and
weather factors and bicycle ridership based on two years of automated
bicycle counts at a Seattle location. In so doing, we fitted a
negative binomial model, and then estimated quantities of interest
using counterfactual simulation. Our findings confirm the significance
of season (+), temperature (+), precipitation (-), as well as holidays
(-), day of the week (+ for Mon-Sat, relative to Sun.), and an overall
trend (+). This paper improves on prior work by demonstrating the use
of the negative binomial instead of a Poisson model, which is
appropriate given the potential for overdispersion as observed in
these data. In addition to validating the significance of factors
identified from the literature, this paper contributes
methodologically through its intuitive visualization of effect sizes
to non-statistical audiences. We believe that the combination of model
type and counterfactual simulation / visualization presented here
reflect a reasonable compromise between model complexity and
intepretability. Results such as these can aid policymakers and
planners both in understanding bicycle travel demand elasticities, and
to guide interventions aimed at increasing rates of
bicycling. Finally, the methods presented here are fully reproducible
and invite adaptation to other locations.

\vspace*{5pc}
\par\noindent\textit{Keywords}: Bicycle counts; negative binomial; season; weather; automated; counterfactual simulation; visualization

\clearpage

\chapter*{Introduction}
%
% 1. Explain what we set out to do (i.e. understand influence of
% seasonal and weather factors on bicycling) 2. explain relevance to
% planners (e.g. understanding how much seasonal variability to expect
% for service provision, understand methodologically how to use data
% collected from counters) 3. ??? other things I'm forgetting right
% now.
%

@@

This article contributes to that understanding by focusing on the
substantive effects of weather and seasonality on the number of
bicyclists observed on a given day. Specifically, we hypothesize that
factors associated with inclement weather (esp. low temperature and
precipitation) will be inversely correlated with bicycle counts. We
similarly expect to see an additional decrease in counts associated
with both physical and social / institutional dimensions of
season. That is, in seasons where the days get shorter, we expect
fewer cyclists due to a kind of decision making inertia. We also
expect fewer cyclists outside the ``school season'' as we believe that
a large number of bicycling trips are associated with commuting to
schools that, unlike most other institutions have clear annual cycles.

@@

\chapter*{Literature Review}
% 
% Identify where people have talked about weather and seasonal factors
% influencing counts previously. Describe their methods together with any
% limitations, as well as their substantive findings---especially what
% variables they considered. Normally should conduct this first, but
% our driver here is a class assignment! We should fill this in and
% adjust our analysis accordingly before attempting to go to press.
%

@@

Research has found the variability for counts has a positive
association with high temperature and low precipitation. \parencite{Niemeier:1996aa,Parkin:2008aa}. Meanwhile, as suggested by
\textcite{Lewin:2011aa} and \textcite{Thomas:2009aa}, the effects of
precipitation and temperature on bicycle volumes are nonlinear. For
example, bicycle traffic can decrease in both very cold and very hot
weather as noted by \textcite{Richardson:2000aa}. Apart from the usual
temperature and rain variables, \textcite{Miranda-Moreno:2011aa} finds
humidity and additional precipitation variables including the presence
of rain in the morning and/or during the previous three hours to be
significant too. Other comparative studies are also available where
bicycle counts are conducted in different cities, and different
sensitives to weather are examined \parencite{Rose:2011aa}. As for
longitudinal studies, \textcite{Niemeier:1996aa} finds increased variability
for counts conducted in the later months of the year.
\textcite{Jones:2010aa} conclude that morning peak hours from 6 AM to 9 PM
accounts for a consistent 95\% of the total bicycle volumes by hourly
count data. The simple linear regression model has been used in
several applications \parencite{Jones:2008aa,Jones:2010aa}. Other modeling approaches include \textcite{Miranda-Moreno:2011aa} which develops a count
model and \textcite{Thomas:2009aa} which develops a time-series model.
\textcite{Niemeier:1996aa} also uses a Poisson model to statistically
confirm  many of the factors thought to influence cyclists. The
work by \textcite{Gallop:2012aa} adopts a similar time-series approach
while incorporating an autoregressive integrated moving average
(ARIMA) analysis. A summary of findings from the literature are
presented in Table \ref{tb:lit}.

These literature together suggest an opportunity for further model
development base around long-term automated counts utilizing
appropriate statistical methods. How seasonal factors influence
bicycle flow needs to be examined in data that last more than a year.
One limitation present in much of the past literature is that few
discuss goodness of fit of their modeling. Further a model that can
better describe and forecast the bicycle count in longitudinal form is
necessary to be developed. Models for count data with better
estimation methods offer some promise.

\begin{table}[htdp]
\begin{center}
\begin{scriptsize}
\caption{Summary of findings from literature review}
\begin{tabularx}{1.0\textwidth}{>{\raggedright\arraybackslash}p{2.7cm}>{\raggedright\arraybackslash}X>{\raggedright\arraybackslash}X}
  \toprule
  Source & Variable(s) identified & Methods \\
  \midrule
  \cite{Fields:2012aa} & Average daily temperature; Total weekly precipitation & Identify patterns through scatter plots; No explicit model is established. \\
\\
  \cite{Gallop:2012aa} & Temperature, Relative humidity, wind speed, visibility, fog, precipitation & Use ARIMA to account for serial correlation patterns \\
\\
  \cite{Griswold:2011aa} & Nearby population and employment density, proximity to downtown/freeway, age, education level, income, etc. & Log linear ordinary least squares regression is used to estimate a bicycle count model \\
\\
  \cite{Helbich:2014aa} & Daily maiximum air temperature, daily average wind speed, daily precipitation & Place-specific associations of weather conditions are explored through geographically weighted logit models \\
\\
  \cite{Hunt:2007aa} & Descriptive variables indicating lane use, secured parking, level of experience, etc. & Logistic model of cycling-related choices \\
\\
  \cite{Jones:2010aa} & Length of bicycle network, employment density, population density & Standard ordinary least squares regression \\
\\
  \cite{Lewin:2011aa} & Max temp, rain flag, snow flag, weekend flag, over 90 flag & Standard linear regression model \\
\\
  \cite{McCahil:2008aa} & logarithmic choice measure, population density, worker density & A new space syntax theory is used to evaluate and predict the bicycle volume throughout a network \\
\\
  \cite{Miranda-Moreno:2011aa} & Temperature, percent humidity, rain presence, rain presence in prev. 3hrs, warm \& humid, morning rain & Both log-linear model and negative binomial model are tested \\
\\
  \cite{Niemeier:1996aa} & Morning flag, rain flag, high temp flag, location variable, season variable & A Poisson count model is assumed and fitted \\
\\
  \cite{Nosal:2014aa} & Temperature, percent humidity, rain flag, rain prev. 3hrs, am rain, pm rain & The relationship is analyzed using a log-linear regression model \\
\\
  \cite{Parkin:2008aa} & Gender, car ownership, hilliness, off-road routes proportion & A logistic regression model of relevant socio-economic and physical variables is estimated. \\
\\
  \cite{Pinjari:2009aa} & Household density, employment density, fraction of commercial land area, demographic factors including proportion of population that are seniors and proportion of population by race & The model system takes the form of a joint mixed Multinomial Logit–Multiple Discrete-Continuous Extreme Value (MNL-MDCEV) structure \\
\\
  \cite{Rose:2011aa} & Temperature, rainfall, holiday flag, school season flag, day of the week & Weather and other effects examined using an aggregate model of daily ridership \\
\\
  \cite{Thomas:2009aa} & Temperature, sunshine, precipitation, wind force, cycle path use & A bi-level structure is developed with the upper level being a log-linear model and the lower level being a linear model \\
  \bottomrule
\label{tb:lit}
\end{tabularx}
\end{scriptsize}
\end{center}
\end{table}

\FloatBarrier

\chapter*{Methodology}

In order to discern the relationship between bicycle counts and
several identified weather, seasonal, and temporal factors, we
developed a statistical model that attempts to predict daily bicycle
counts from these other factors. This section describes the methods
and procedures we used to collect and process the estimation dataset,
the rationale for our selection of variables, our chosen model type,
and our model estimation procedures.

\FloatBarrier
\section*{Data collection, processing, description}
%
% Data collection: study site, descriptives of sample, collection from
% forecast.io and seattle.gov, not in that order
%

The data that we attempt to predict were collected at Seattle's
Fremont Bridge and cover a period of two years spanning from October
31, 2012 to October 30, 2014. The Fremont Bridge captures a
substantial amount of bicycle traffic due to its status as one of only
five facilities that carry bicyclists across the canal separating the
northern and southern halves of Seattle.

@@

Figure \ref{fg:descriptives} provides a visual summary of the processed
counts data. Some apparent outliers are visible at the rightmost
portion of the histogram. The two highest counts occurred the Monday
and Tuesday preceding the beginning of National Bike to Work
Month. And the third highest count occurred on National Bike to Work
Day.

\begin{figure}[ht!]
  \centering
  \subbottom[Frequency of observed counts\label{fig:top}]%
    {\includegraphics[width=0.5\textwidth]{../fig/dp1-crop}}\hfill
  \subbottom[Timeseries of counts\label{fig:bottom}]%
    {\includegraphics[width=0.5\textwidth]{../fig/dp2-crop}}
  \caption{Descriptive visualization of bicycle counts dataset.}
  \label{fg:descriptives}
\end{figure}

% \begin{figure}[h!]
%   \centering
%   \includegraphics[width=0.5\textwidth]{../fig/dp2}
%   \caption{Timeseries plot of bicycle counts}
%   \label{fg:timeseries}
% \end{figure}

\FloatBarrier
\section*{Variable selection}
% 
% Data analysis: specification of model, choice of model (nb v pois),
% description of counterfactual simulation using the estimated model,
% software used.
%

Out of the set of independent variables retrieved from Forecast.io and
the University of Washington, we selected a subset that we felt best
reflected our specific research questions. The final set of variables
included in our model include those presented and summarized in Table
\ref{tb:variables}. Further explanation and justification of these
variable choices follow.

\begin{table}[ht!]
\begin{center}
\begin{scriptsize}
\caption{Variables included in model specification}
\begin{tabularx}{0.7\textwidth}{>{\raggedright\arraybackslash}p{1.6cm}>{\raggedright\arraybackslash}X}
\toprule
Variable & Description \\
\midrule
Count* & Number of bicycles per day \\
Daylight  & Length of daylight in hours \\
UW & The University of Washington was in session (TRUE/FALSE) \\
Holiday & The day was a holiday as recognized by UW (TRUE/FALSE) \\
Max temp & Maximum temperature in Fahrenheit for the day \\
Max temp\textsuperscript{2} & \texttt{Max temp} squared \\
Max precip & Daily maximum inches of precipitation in any hour \\
Sat--Fri & Day of the week dummy variable (TRUE/FALSE, relative to Sunday) \\
Day \# & Sequentially numbered day of study \\
\bottomrule
\multicolumn{2}{r}{* Dependent variable}
\end{tabularx}

\label{tb:variables}
\end{scriptsize}
\end{center}
\end{table}

@@

Daily maximum temperature, measured in Fahrenheit, was chosen to
represent temperature (rather than, for example, substituting or
adding daily minimum temperature) in part to retain simplicity in the
model, in part because there is relatively little daily temperature
variation in Seattle due to the moderating effect of large water
bodies, and in part because maximum temperature better reflects the
conditions during daylight hours when most bicycle trips would occur.
This simplification may not be warranted for other locations that
experience greater temperature variation than Seattle. In addition to
daily maximum temperature, we also included a daily maximum
temperature squared term in order to identify the potential for a
leveling off of counts in very high temperatures.

@@

\FloatBarrier
\section*{Model estimation and goodness of fit}
% Model choice
A natural model choice for count data is the Poisson model, however we
believed that our count data were overdispersed. Overdispersion, or
contagion between events, violates the mean-variance equivalence
assumption of the Poisson model. While overdispersion would not impact
our parameter estimates, it would result in overly optimistic margins
of error. In order to account for and estimate the amount of
overdispersion present in our data, we chose a negative binomial model
type. Because there were no days observed with zero bicycle counts, we
did not need to resort to zero-inflated models as is often necessary
when dealing with count data.

@@

Additionally, we compared the residuals plots for both the Poisson and
negative binomial models. As seen in figure \ref{fg:respois}, the
residual variance is much lower in the negative binomial model,
further confirming the choice of a negative binomial for these counts.

@@

\begin{figure}[ht!]
  \centering
  \subbottom[Poisson\label{fig:mdleft}]%
    {\includegraphics[width=0.5\textwidth]{../fig/residuals_pois-crop}}\hfill
  \subbottom[Negative binomial\label{fig:mdright}]%
    {\includegraphics[width=0.5\textwidth]{../fig/residuals-crop}}
  \caption{Residual variance of Poisson vs. negative binomial model.}
  \label{fg:respois}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/avpp2-sm}
  \caption{Actual versus predicted values as fit by negative binomial model.}
  \label{fg:avp}
\end{figure}

@@

\chapter*{Results}
% 
% One subsection for each research question
% 
This section presents the results from the statistical model and
accompanying counterfactual simulations as described in the preceding
section. Each of the main research questions of seasonality, weather,
and general trend are addressed here. In addition, additional added
control variables such as day of the week, holidays, and linear trend
are presented.

With the exception of the \texttt{Max temp\textsuperscript{2}} term,
each of the coefficients in our model was statistically significant at
the $p < 0.05$ level. Further, with the exception of the \texttt{Max
  temp\textsuperscript{2}} and \texttt{Saturday} coefficients , all
coefficients were significant at the $p < 0.001$ level. While this is
instructive in confirming the significance of these factors, the
remainder of this section focuses on presenting the substantive effect
of each variable.

\FloatBarrier
\section*{Seasonality}
As discussed previously, we considered two variables to address the
question of seasonality: the first being the number of daylight hours,
and the second being whether or not the University of Washington
(proxying more generally for other educational institutions) was in
session.

@@

\FloatBarrier
\section*{Weather}
For our examination of seasonality, our model represents weather using
three variables. In this case, these are precipitation (measured as
the maximum amount of precipitation falling in any hour of that day),
maximum temperature (measured daily, in Fahrenheit), and maximum
temperature squared.

@@

Temperature, in contrast to precipitation, has a clearly positive
association with increased numbers of bicyclists. Temperature squared
was not significant in our model at the $p < 0.05$ level. We were
somewhat surprised to not see evidence for a leveling off in counts at
very high temperatures---likely due to Seattle's moderate summer
climate, and the lack of extremely warm days in our dataset.

@@

\FloatBarrier
\section*{General trend in bicycle counts}
Our results (shown in figure \ref{fg:trend}) confirm the presence of a
general trend toward increased numbers of bicycles at this location.
With all else held constant, we would expect to see roughly 300 more
bicycles on days at the end of our study period than at the beginning.
This is consistent with figures reported elsewhere that suggest that
bicycling is increasing in a number of cities including
Seattle \parencite{League-of-American-Bicyclists:aa}.

@@

\FloatBarrier
\section*{Day of week variation}
As discussed previously, day of the week was included due to the
weekly variation in bicycle counts apparent in timeseries plots.
Figure \ref{fg:dayofweek} shows several interesting aspects of these
results. First, we see much higher numbers of bicyclists on weekdays
than on weekends. This strongly suggests that the majority of the
bicycle traffic at this location is for commuter purposes.

@@

\FloatBarrier
\chapter*{Discussion}
This research set out to better understand the relationship between
bicycle counts and weather and season. However, weather and season are
themselves comprised by a number of constituent elements. We believe
this research has chosen defensible representations of weather and
season. However, we readily concede that there exist other reasonable
representations of season and weather.

@@

One variable missing from our analysis that we would have liked to
include is cloud cover. This variable would have allowed us to account
for the possibility of brilliant, sunny, but cold winter days, as well
as warm, dry, but gloomy summer days. Unfortunately, the dataset
downloaded from Forecast.io included a great deal of missing values
for this variable, and thus future work might include identification
of a more suitable source of these data.

Another limitation of this research comes from our sample of bicycle
counts. While we do have continuous counts spanning a full two years,
these counts were taken at just one point location. We believe the
Fremont bridge is somewhat representative of a high-volume bicycle
facility in Seattle. However we might, for example, see a decreased
effect size for University of Washington session status if we looked
at counts further from the university. The City of Seattle has begun
collecting and releasing data from other counters, which could be
included in a future version of this analysis.

Finally, as discussed in the methodology section, we believe the
choice of the negative binomial model to be a better choice than the
Poisson, which had been used in some previous studies. A limitation in
this model, however, is in not accounting for timeseries
autocorrelation. We think it highly likely that the decision to
bicycle on one day might influence the decision to bicycle on the
next. A future direction would be to consider models, such as ARIMA,
that more explicitly account for time, though that decision would
represent a potentially undesirable increase in model complexity.

\chapter*{Conclusions}
This research set out to help clarify the relationship between weather
and seasonal factors and bicycle counts in Seattle. In order to
achieve this, we developed a negative binomial model to predict
bicycle counts based on temperature, precipitation, day length,
university in-session status, day of the week, and a general linear
time trend.

For each term included in our model except temperature squared, we
found statistically significant effects. More importantly, through the
use of counterfactual simulation, we estimated what we deem to be
substantial effect sizes associated with each predictor variable.

This article contributes to the existing literature by demonstrating
the use of an appropriately chosen negative binomial model for bicycle
counts. It provides an additional methodological contribution in
illustrating counterfactual simulation and visualization to create
more compelling and intuitive results summaries for
non-statisticians. Finally, the results presented here were generated
from data collected over a relatively long study period of two years.
